\thetitle

### 2008-01-22 models.tex
Finished updating the methodology.

\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited src/dst distributions

@@

\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals drawn from fitted Exponential distributions

>>>>

\raisebox{1.5pt}{$\centerdot$} nodes assigned as flow end-points by sampling from inherited distributions

@@

\raisebox{1.5pt}{$\centerdot$} packet sizes and intervals sampled from fitted Exponential distributions

### 2008-01-22 paper.tex
Finished updating the methodology.

We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network. The standard na\"ive model uses the \class{Uniform} sub-model at all levels. The models proposed by {\FHC} are implemented using the \class{Parametric} sub-model at every level. The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.

@@

\begin{table*}[t]
\begin{center}
\small
\input{models.tex}
\footnotesize\\[0.5em]
%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
\end{center}
\vspace{-0.75em}
\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
\label{tab:models}
\vspace{-1.75em}
\end{table*}

@@

Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be played either by an actual abstract class or by an interface. In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if inheritance is desired, the base class can be made abstract by defining stub methods that throw informative exceptions if called.} 
Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work. Other traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper. Concrete classes implement the behavior of specific traffic models at each level. A complete synthetic workload is instantiated as follows:
\begin{enumerate}
\item create a \class{Network} object of the appropriate class;
\item the top-level \class{Network} object in turn instantiates a collection of \class{Node} objects and associated \class{Flow} objects;
\item each \class{Flow} object generates a stream of \class{Packet} objects.
\end{enumerate}
The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation. Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).

@@

For each of the \class{Node}, \class{Flow} and \class{Packet} abstract classes, we create define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric} and \class{Uniform}. The \class{Trace} class at each level takes its properties directly from the behavior of the original traffic. The \class{Nonparametric} and \class{Parametric} classes use nonparametric models of marginal behavior at each level to

We separate traffic generation into three levels of behavior: network, flow and packet. The {network} level determines properties pertaining to nodes and their interaction. For example, it determines which nodes communicate with each other, and how frequently. This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models. The next level of behavior determines overall properties of flows. This includes each flow's start time, duration, packet count and total data volume. Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission. Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine. Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.

@@

For each level of behavior we define a spectrum of models. The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly. The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets. Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments. Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity. At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).

\begin{figure*}[tb]
\begin{center}
\subfloat[trace node behavior]
{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[uniform node behavior]
{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[trace flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[uniform flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[packet behavior]
{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
\caption{Examples illustrating trace versus uniform behaviors a the node, flow and packet levels. Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway). Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Figure \ref{fig:packet-behavior} compares \caps{CBR} packet behavior with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
\label{fig:traffic-models}
\end{center}
\vspace{-2em}
\end{figure*}

@@

The three levels of behavior are orthogonal and can be varied almost independently. The exception to their independence is that the trace-based packet behavior model can only be used when the flow behavior model is also trace-based. Once flow behavior is decoupled from the actual trace, there is no natural way to preserve packet behavior. This eliminates three combinations of models and leaves nine viable behaviors, abbreviated by the first letters of their flow topology, flow behavior, and packet behavior models: \caps{TTT} (fully trace-based), \caps{TTU}, \caps{TUU}, \caps{STT}, \caps{STU}, \caps{SUU} (entirely synthetic, sink topology), \caps{UTT}, \caps{UTU}, \caps{UUU} (entirely synthetic, uniform topology).

@@

We have selected a variety of performance metrics at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), routing queue drop rate.
\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate.
% TODO: are we actually using RTS/CTS?
\end{enumerate}
These metrics are commonly used to evaluate the performance of new wireless protocols. For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

We use the Qualnet wireless network simulator to perform our experiments. We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

>>>>

We define a range of sub-models at each level of the \caps{TOM}, with decreasing similarity to the original trace behavior: \class{Trace} (reproduce original behavior), \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The control model uses the Trace sub-model at every level, exactly reproducing the behavior recorded from the original network. The standard na\"ive model uses the \class{Uniform} sub-model at all levels. The models proposed by {\FHC} are implemented by applying the \class{Parametric} sub-models at all levels. The \class{Nonparametric} sub-models provide a measure of how much error is introduced by parametric approximation and parameter estimation.

@@

\begin{table*}[t]
\begin{center}
\small
\input{models.tex}
\footnotesize\\[0.5em]
%\raisebox{1pt}{*}Parametric models fitted to inherited empirical distributions using maximum likelihood estimation ({\scriptsize{MLE}}).
\end{center}
\vspace{-0.75em}
\caption{Details of how the sub-models at each behavioral level of the {\footnotesize{TOM}} inherit, override and delegate properties.}
\label{tab:models}
\vspace{-1em}
\end{table*}

@@

% TODO: fix-up client-server vs. source-destination terminology.

Abstract base classes define the properties at each level that traffic models must implement.\footnote{In Java or C\#, this role may be served either by an actual abstract class or by an interface. In duck-typed languages such as Ruby or Python, no actual abstract class need exist, since type is implicitly defined by usage; if abstract classes are desired, they can be simulated by creating base classes with stub methods that throw informative exceptions upon being called.} 
Table~\ref{tab:properties} enumerates all the properties exposed by traffic models used in this work. Future traffic models may necessitate exposing additional properties, but this list suffices to implement the models used in this paper. Concrete subclasses implement the behavior of specific traffic models at each level. A complete synthetic workload is instantiated as follows:
\begin{enumerate}
\item create a \class{Network} object of the appropriate subclass;
\item the top-level \class{Network} object in turn creates a collection of \class{Node} objects and associated \class{Flow} objects;
\item each \class{Flow} object generates a stream of \class{Packet} objects.
\end{enumerate}
The hierarchy of objects may be used to directly drive a simulation, or may generate configuration files that in turn drive a simulation. Our implementation does the latter, generating a collection of Qualnet configuration files and trace files specific to each workload instantiation (see \Section{simulations}).

@@

For each of the \class{Node}, \class{Flow} and \class{Packet} abstract superclasses, we define four concrete subclasses: \class{Trace}, \class{Nonparametric}, \class{Parametric}, and \class{Uniform}. The \class{Trace} classes take properties of objects directly from the behavior of the corresponding object in the original trace. The \class{Nonparametric} and \class{Parametric} classes use models of the corresponding type to randomly generate the properties of objects. The \class{Uniform} classes generate identical objects with average properties. The specific details of how each the concrete classes at each level are defined are specified in Table~\ref{tab:models}. The parameters for the fitted models used in the \class{Parametric} subclasses are estimated from the corresponding nonparametric distribution using maximum likelihood estimation. Figure~\ref{fig:trace-vs-uniform} shows visual depictions comparing trace versus uniform behavior at various levels in example scenarios.

While the modularity of the Traffic Object Model allows us to mix and match sub-models, not all possible combinations are valid. In particular, note that the \class{Trace} classes are tied to a specific object from the underlying trace, whereas for other classes, there is no such correspondence to any specific object. Consequently, once a non-trace model has been used, trace models may not be used at the lower levels. This still leaves 40 valid composite combinations of sub-models; we only evaluate twelve of these. Composite models are abbreviated by the initials of their constituent sub-models. The fully trace-base model, \class{TTT}, serves as an experimental control. We consider \class{TTN}, \class{TTP}, and \class{TTU} in order to compare models of packet behavior. We evaluate different flow models, using the best possible model at the packet level: \class{TNN}, \class{TPN}, \class{TUN}. Similarly, we compare models for node behavior: \class{NNN}, \class{PNN}, \class{UNN}. Finally, we use the na\"ive uniform model, \class{UUU}, and the composite parametric model, \class{PPP}, proposed by {\FHC}~\cite{Hernandez06:wlan-traffic}, as external points of comparison.

\begin{figure*}[tb]
\begin{center}
\subfloat[trace node behavior]
{\includegraphics[width=1.392in]{flow_topology_trace}\label{fig:flow-topology-trace}}
\subfloat[uniform node behavior]
{\includegraphics[width=1.392in]{flow_topology_uniform}\label{fig:flow-topology-uniform}}
\subfloat[trace flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_trace}\label{fig:flow-behavior-trace}}
\subfloat[uniform flow behavior]
{\includegraphics[width=1.392in]{flow_behavior_uniform}\label{fig:flow-behavior-uniform}}
\subfloat[packet behavior]
{\includegraphics[width=1.392in]{flow_packet_behavior}\label{fig:packet-behavior}}
\caption{Examples illustrating trace versus uniform behaviors a the node, flow and packet levels. Figures \ref{fig:flow-topology-trace} and \ref{fig:flow-topology-uniform} show example node behaviors. The width of each line is proportional to the logarithm of the number of flows between the nodes it connects (zero is the Internet gateway). Uniform and trace flow behavior examples are plotted in Figures \ref{fig:flow-behavior-trace} and \ref{fig:flow-behavior-uniform}. The time axis indicates when flows start and end; the width of each flow line is proportional to the logarithm of its data rate. Figure \ref{fig:packet-behavior} compares uniform (i.e. {\footnotesize{CBR}}) packet behavior with the trace of an actual flow. In the uniform model, the cumulative data sent increases smoothly over time, whereas in the actual packet trace, the transmissions are variable both in size and in inter-packet interval, leading to a ``lumpy'' cumulative data plot.}
\label{fig:trace-vs-uniform}
\end{center}
\vspace{-2em}
\end{figure*}

%We separate traffic generation into three levels of behavior: network, flow and packet. The {network} level determines properties pertaining to nodes and their interaction. For example, it determines which nodes communicate with each other, and how frequently. This level can be viewed as assigning network nodes to the source and destination roles of flows, as well as determining per-node properties which affect lower level models. The next level of behavior determines overall properties of flows. This includes each flow's start time, duration, packet count and total data volume. Finally, the packet level of behavior determines the individual sizes of packets and the durations of intervals between their transmission. Despite the concreteness of the examples mentioned here, there are also less obvious, abstract properties that models at the network and flow levels determine. Flow models, for example, dictate the distribution of each flow's packet sizes, a property which can be used to determine individual packet sizes, but which is not itself specific to any particular packet.

@@

%For each level of behavior we define a spectrum of models. The gold standard of realism at each level is provided by the ``trace'' model for that level, which replicates the node, flow or packet behavior of the original trace exactly. The most rudimentary model at each level is the ``uniform'' model, which treats all entities identically---be they nodes, flows or packets. Despite being so simplistic (or perhaps because of it), the uniform behavior models are by far the most commonly for workload generation in experiments. Between these extremes on the spectrum of realism, we consider one or two models with intermediate degrees of complexity and fidelity. At the packet level, we define the ``sample'' model, which is in essence a per-flow variable bit-rate (\caps{VBR}) model: each packet's size is determined by randomly sampling a value from the packet size distribution given by the flow model; the interval until the next packet is then sampled from the given interval distribution.\footnote{Note that the uniform packet model (\textit{i.e.} {\scriptsize{CBR}}), can be considered to be a special case of the sample model where the size and time distributions are both constant.} The intermediate model at the flow level is designated as ``proportional'': the distributions of packet sizes and inter-packet intervals are taken from the trace and passed down to the packet model (which may use the actual distribution, via sampling, or simply use the mean size and intervals, to generate \caps{CBR} traffic).

@@

%The three levels of behavior are orthogonal and can be varied almost independently. The exception to their independence is that the trace-based packet behavior model can only be used when the flow behavior model is also trace-based. Once flow behavior is decoupled from the actual trace, there is no natural way to preserve packet behavior. This eliminates three combinations of models and leaves nine viable behaviors, abbreviated by the first letters of their flow topology, flow behavior, and packet behavior models: \caps{TTT} (fully trace-based), \caps{TTU}, \caps{TUU}, \caps{STT}, \caps{STU}, \caps{SUU} (entirely synthetic, sink topology), \caps{UTT}, \caps{UTU}, \caps{UUU} (entirely synthetic, uniform topology).

@@

We have selected seven performance metrics at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, average jitter, total received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), routing queue drop rate.
\item \textbf{Link (\caps{MAC}):} control overhead (\caps{RTS/CTS/ACK}), packet retransmission rate.
% TODO: are we actually using RTS/CTS?
\end{enumerate}
These metrics are commonly used to evaluate the performance of wireless protocols.%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments. We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

We have modified the standard Qualnet distribution in the two significant ways. First, In our previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}, we approximated full-duplex \caps{TCP} with a pair of \caps{UDP} flows. For this work, we have implemented a new Qualnet application driver that allows trace-driven full-duplex \caps{TCP} flows. This allows us to accurately reproduce the full dynamics of \caps{TCP} feedback. Moreover, raw \caps{IP} and \caps{UDP} flows are implemented using the same code, guaranteeing that all traffic is simulated uniformly and that all performance metrics are aggregated in the same manner. The second modification is that we have re-instrumented Qualnet to collect \caps{IP} performance metrics in strict adherence to the recommendations of the \caps{IETF} working group recommendations~\cite{rfc:ip-metrics,rfc:jitter}.
